---
title: ""
date: '2018-03-07'
---

```{r "setup", include=F,echo=F, warning=F, error=F}
library(tweenr); library(gganimate); 
library(tidyverse); library(scholar)
library(rio); library(knitr)
options(knitr.kable.NA = '')
CV = import_list("../../../Dropbox/Privat/Jobbsøking/CV-attester-etc/CV.xlsx")
```

### In short
I am a cognitive neuroscientist based in Oslo, Norway, and works at the [Center for Lifespan Changes in Brain and Cognition](www.oslobrains.no). I'm passionate about cognitive neuroscience, using R and teaching the world the wonders of R. Doing all I can to improve gender diversity in the R-community though [RLadies-Oslo](https://twitter.com/RLadies_Oslo).


### Education  
```{r "Education", echo=F, results='asis'}
CV$EDU %>% 
  kable("markdown", booktabs = T)

```
### Work  
```{r "Work", echo=F}
CV$WORK %>% 
  kable("markdown", booktabs = T)
```
### Commissions  
```{r "Commissions", echo=F}
CV$COMMISSION %>% 
  kable("markdown", booktabs = T)
```


### Publications  


```{r "pubPlot", echo=FALSE, warning=FALSE, error=FALSE, dev.args = list(bg = 'transparent'), aniopts=c(interval = 0.8, loop=1)}
pp = get_citation_history("7NkxgWQAAAAJ") %>% 
  mutate(cumulative=cumsum(cites)) %>% 
  gather(metric, val, -1) %>% 
  mutate(ease = 'cubic-in-out') %>% 
  tween_elements("year", "metric", "ease", nframes = nrow(.)/2) %>% 
  ggplot(aes(x=year, y=val, group=.group, fill=val, color=.group, frame=.frame,  cumulative = TRUE)) + 
  geom_line(lineend="round") + 
  geom_point(show.legend = F) +
  annotate("text",x=-Inf,y=Inf,vjust=1.5,hjust=-0.05,size=3,colour='gray',
           label=format(Sys.time(), "%d-%m-%Y %H:%M %Z")) + #add timestamp for the update
  labs(y="Googe citations", x="Year") +
  theme(panel.grid.major = element_line(color="#4E67691F"),
        panel.background =  element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank(), 
        legend.background = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key = element_blank(),
        axis.line = element_line(),
        text = element_text(color="#4E6769"),
        line = element_line(color="#4E6769")) +
  scale_color_manual(values=c("#2c3b3b", "#AD957F"))

#gganimate(pp, title_frame = FALSE)
pp
```


```{r "pubList", results='asis', warning=FALSE, echo=FALSE}
pubs = get_publications("7NkxgWQAAAAJ") %>% 
  filter(journal != "") %>% 
  mutate(cid     = ifelse(is.na(cid), NA, paste0("https://scholar.google.no/scholar?oi=bibs&hl=en&cluster=", cid))) %>% 
  mutate(author  = gsub("Mowinckel", "**Mowinckel**", author), 
         journal = ifelse(is.na(cid), paste0("_", journal,"_"), paste0("[_", journal,"_](", cid, ")")), 
         number  = paste0("_",number,"_"),
         year    = paste0("(",year,")")) %>% 
  arrange(desc(year)) %>% 
  transmute(pub=paste(author, year, title, journal, number, collapse="<br><br>")) 
cat(pubs[1,])


```

